---
title: "Chapter 9: Linear regression"
author: "Jesse Mu"
date: "November 2, 2016"
output:
  html_document:
    highlight: pygments
    toc: yes
    toc_float: yes
---

<!-- Setup -->

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: { 
      equationNumbers: {
            autoNumber: "all"
      } 
  }
});
</script>

```{r echo=FALSE, message=FALSE}
knitr::opts_chunk$set(fig.align = 'center', message = FALSE)
library(knitr)
library(ggplot2)
library(cowplot)
library(reshape)
```

<!-- Begin writing -->

# Exercises

## 9.1

```{r}
swim = read.table(url('http://www.stat.washington.edu/people/pdhoff/Book/Data/hwdata/swim.dat'))
```

To specify our prior, we let the prior expectation of our $y$-intercept to be 
23, and we let the prior expectation of the effect of training week by week to
be 0, so $\boldsymbol{\beta} = (23, 0)^T$.

We expect no covariance with the $\beta$ coefficients, but we do have
uncertainty about our initial $\beta$ estimates. Specifically, to let 95% of our
uncertainty of the $y$-intercept to fall in $[22, 24]$, we let $\Sigma_{0(1, 1)} = 1/4$ (so that $\pm$ 2 standard deviations is $\pm$ 1).
We also expect that training has a relatively mild effect on time, so we let
$\Sigma_{0(2, 2)} = 0.1$ which is just an arbitrarily chosen small variance.

For our expectation of the variability of measurements, let's similarly set 
$\sigma^2_0 = 1/4$ and only lightly center this prior with $\nu_0 = 1$.

```{r}
library(MASS) # mvrnorm
library(dplyr)
S = 5000
X = cbind(rep(1, 6), seq(1, 11, by = 2))
n = dim(X)[1]
p = dim(X)[2]

# Prior
beta0 = c(23, 0)
sigma0 = rbind(c(0.25, 0), c(0, 0.1))
nu0 = 1
s20 = 0.25

set.seed(1)

inv = solve

# For each swimmer, run linear regression gibbs sampling and obtain a posterior
# predictive distribution
swim_pred = apply(swim, MARGIN = 1, function(y) {
  
  # Store samples
  BETA = matrix(nrow = S, ncol = length(beta0))
  SIGMA = numeric(S)
  
  # Starting values - just use prior values?
  beta = c(23, 0)
  s2 = 0.7^2
  
  # Gibbs sampling algorithm from 9.2.1
  for (s in 1:S) {
    # 1a) Compute V and m
    V = inv(inv(sigma0) + (t(X) %*% X) / s2)
    m = V %*% (inv(sigma0) %*% beta0 + (t(X) %*% y) / s2)
    
    # 1b) sample beta
    beta = mvrnorm(1, m, V)
    
    # 2a) Compute SSR(beta) (specific formula from 9.1)
    ssr = (t(y) %*% y) - (2 * t(beta) %*% t(X) %*% y) + (t(beta) %*% t(X) %*% X %*% beta)
    
    # 2b) sample s2
    s2 = 1 / rgamma(1, (nu0 + n) / 2, (nu0 * s20 + ssr) / 2)

    BETA[s, ] = beta
    SIGMA[s] = s2
  }
  
  # Now sample posterior predictive - two weeks later
  xpred = c(1, 13)
  YPRED = rnorm(S, BETA %*% xpred, sqrt(SIGMA))
  
  YPRED
})

```

### b

```{r}
fastest_times = apply(swim_pred, MARGIN = 1, FUN = which.min)
table(fastest_times) / length(fastest_times)
```

We notice that with our posterior predictive dataset, swimmer 1 is the fastest 
about 65% of the time by week 13, so we recommend that swimmer 1 race.

## 9.3

```{r}
crime = read.table(url('http://www.stat.washington.edu/people/pdhoff/Book/Data/hwdata/crime.dat'), header = TRUE)
```

### a

```{r}
y = crime$y
X = crime %>% select(-y) %>% as.matrix

n = dim(X)[1]
p = dim(X)[2]

g = n
nu0 = 2
s20 = 1

S = 1000

Hg = (g / (g + 1)) * X %*% inv(t(X) %*% X) %*% t(X)
SSRg = t(y) %*% (diag(1, nrow = n) - Hg) %*% y

s2 = 1 / rgamma(S, (nu0 + n) / 2, (nu0 * s20 + SSRg) / 2)
Vb = g * inv(t(X) %*% X) / (g + 1)
Eb = Vb %*% t(X) %*% y

E = matrix(rnorm(S * p, 0, sqrt(s2)), S, p)
beta = t(t(E %*% chol(Vb)) + c(Eb))
```


```{r}
library(tidyr)

signif = apply(beta, MARGIN = 2, FUN = quantile, probs = c(0.025, 0.5, 0.975)) %>%
  apply(MARGIN = 2, FUN = function(y) !(y[1] < 0 && 0 < y[3]))

beta_df = as.data.frame(beta) %>%
  gather(key = 'variable', val = 'coefficient') %>%
  mutate(signif = signif[variable])

ggplot(beta_df, aes(x = variable, y = coefficient, color = signif)) +
  stat_summary(fun.y = mean, fun.ymin = function(y) quantile(y, probs = c(0.025)), fun.ymax = function(y) quantile(y, probs = c(0.975))) +
  geom_hline(yintercept = 0, lty = 2)
```

Looks like Ed (mean years of schooling), Ineq (Income inequality), M (percentage
of males aged 14-24), Prob (probability of imprisonment), and U2 (unemployment
rate of urban males 35-39).

### b

```{r}
y = crime$y
X = crime %>% select(-y) %>% as.matrix
set.seed(1) # Reproducible!
train_i = sample.int(length(y), size = round(length(y) / 2), replace = FALSE)
ytr = y[train_i]
Xtr = X[train_i, ]
yte = y[-train_i]
Xte = X[-train_i, ]
```

#### i

```{r}
# From 9.1
beta_ols = inv(t(Xtr) %*% Xtr) %*% t(Xtr) %*% ytr
beta_ols

y_ols = Xte %*% beta_ols

ols_df = data.frame(
  observed = yte,
  predicted = y_ols
)

ggplot(ols_df, aes(x = observed, y = predicted)) +
  geom_point() +
  geom_smooth(method = 'lm')

pred_error = sum((yte - y_ols)^2) / length(yte)
pred_error
```

#### ii

```{r}
y = ytr
X = Xtr

n = dim(X)[1]
p = dim(X)[2]

g = n
nu0 = 2
s20 = 1

S = 1000

Hg = (g / (g + 1)) * X %*% inv(t(X) %*% X) %*% t(X)
SSRg = t(y) %*% (diag(1, nrow = n) - Hg) %*% y

s2 = 1 / rgamma(S, (nu0 + n) / 2, (nu0 * s20 + SSRg) / 2)
Vb = g * inv(t(X) %*% X) / (g + 1)
Eb = Vb %*% t(X) %*% y

E = matrix(rnorm(S * p, 0, sqrt(s2)), S, p)
beta = t(t(E %*% chol(Vb)) + c(Eb))

beta_bayes = as.matrix(colMeans(beta))

y_bayes = Xte %*% beta_bayes

bayes_df = data.frame(
  observed = yte,
  predicted = y_bayes
)

ggplot(ols_df, aes(x = observed, y = predicted)) +
  geom_point() +
  geom_smooth(method = 'lm')

pred_error = sum((yte - y_bayes)^2) / length(yte)
pred_error
```

At least when the seed is 1, there doesn't appear to be a major difference 
between the prediction errors.

### c

```{r}
# Unnecessary code repeating here. Clean this up later w/ functions
N = 100
set.seed(1)
pred_errors = t(sapply(1:N, function(i) {
  y = crime$y
  X = crime %>% select(-y) %>% as.matrix
  train_i = sample.int(length(y), size = round(length(y) / 2), replace = FALSE)
  ytr = y[train_i]
  Xtr = X[train_i, ]
  yte = y[-train_i]
  Xte = X[-train_i, ]
  
  # OLS
  beta_ols = inv(t(Xtr) %*% Xtr) %*% t(Xtr) %*% ytr
  beta_ols
  
  y_ols = Xte %*% beta_ols
  
  pred_error_ols = sum((yte - y_ols)^2) / length(yte)
  
  # Bayes
  y = ytr
  X = Xtr
  
  n = dim(X)[1]
  p = dim(X)[2]
  
  g = n
  nu0 = 2
  s20 = 1
  
  S = 1000
  
  Hg = (g / (g + 1)) * X %*% inv(t(X) %*% X) %*% t(X)
  SSRg = t(y) %*% (diag(1, nrow = n) - Hg) %*% y
  
  s2 = 1 / rgamma(S, (nu0 + n) / 2, (nu0 * s20 + SSRg) / 2)
  Vb = g * inv(t(X) %*% X) / (g + 1)
  Eb = Vb %*% t(X) %*% y
  
  E = matrix(rnorm(S * p, 0, sqrt(s2)), S, p)
  beta = t(t(E %*% chol(Vb)) + c(Eb))
  
  beta_bayes = as.matrix(colMeans(beta))
  
  y_bayes = Xte %*% beta_bayes
  
  pred_error_bayes = sum((yte - y_bayes)^2) / length(yte)

  c(pred_error_ols, pred_error_bayes)
})) %>% as.data.frame
colnames(pred_errors) = c('ols', 'bayes')
```

Here's a plot of the density of $\text{err}_{\text{Bayes}} -
\text{err}_{\text{ols}}$. If this is less than 0, then the Bayes estimator did
better than the OLS estimator:

```{r}
pred_diff = pred_errors %>% transmute(`bayes - ols` = bayes - ols)

ggplot(pred_diff, aes(x = `bayes - ols`)) +
  geom_density() +
  geom_vline(xintercept = 0, lty = 2)
```

Lastly,

```{r}
mean(pred_errors$bayes < pred_errors$ols)
```

For `N` samples, `r mean(pred_errors$bayes < pred_errors$ols) * 100`% of the
time, the predictive error using the Bayes estimators is less than the
predictive error using the OLS estimators. Nice!
